A multi-scale story of the diffusion of a new technology: the web

### 2025-03-12 paper/submission2/tranos2025.qmd
paper v2

---
title: "A multi-scale story of the diffusion of a new technology: the web"
author:
  - name: Emmanouil Tranos
    email: e.tranos@bristol.ac.uk
    affiliations: 
      - id: some-tech
        name: University of Bristol
        department: 
        address: 
        city: 
        state: State
        postal-code: UK
    corresponding: e.tranos@bristol.ac.uk
address:
  - attributes:
        corresponding: true
    note: This is the first author footnote.
abstract: |
    Despite the broad consensus that differences in technological adoption and sophistication can lead to uneven trajectories of local development as new technologies can increase productivity and allow firms to introduce new products and services, economic geographers have shied away from exploring the diffusion of digital technologies across space and time. This paper fills this gap by exposing the roles of different spatial mechanisms in the diffusion of an intangible, digital technology at very granular geographical scales. It does so by employing a novel source of data and machine learning algorithms to model the active engagement with Web during 1996-2012.
keywords: 
  - diffusion
  - technologies
  - Web
  - digital
date: last-modified
bibliography: bibliography
format:
  elsevier-pdf:
    keep-tex: true
    text: |
      \usepackage[demo]{graphicx} % "demo" option just for this example
      \usepackage{subcaption}
    journal:
      name: Journal Name
      formatting: preprint
      model: 3p
      cite-style: authoryear
editor: 
  markdown: 
    wrap: 72
---

@@

Although the above processes have been studied in innovation studies,
the distancing of economic geographers from research questions about the
diffusion of new technologies and innovations resulted to a knowledge gap
regarding the explanatory power of these mechanisms at granular, local
scales [@ding2010modeling]. As discussed in @sec-introduction, such 
spatially explicit knowledge can directly inform local technology
and infrastructure policies. Indeed, while the spatial dimension is
present in the following studies, the geographical level of analysis is
often too coarse to ascertain the role of the above mechanisms at the
local scale. @beardsell1999spatial studied the evolution of the computer
industry in 317 US metropolitan areas during the 1977-1992 period using
employment data. Their analysis indicated that the relative size
distribution holds for urban computer employment and also urban
heterogeneity is essential in explaining this distribution. In a recent
study, @bednarz2020pulled focused on wind turbines and modelled their
spatial diffusion across 402 German regions during 1970-2015. Their key
finding is that local demand than local supply was the main driving
factor. @haller2011determinants employed firm panel data for Ireland
during 2003-2005 and illustrated that when firms are located in a region
or industry with high share of firms having a website, they are more
likely to have a website too. At a global scale @perkins2011internet
explored whether the adoption of previous communication technologies
that is mail, telegrams and telephones were shaped by similar
socioeconomic factors as the Internet. Their results indicated common
patterns regarding the drivers behind the adoption of different
communication technologies. @feldman2015rknowledge focused on inventions
and used patent data to map how the the rDNA technology diffused among
inventors across 366 US metropolitan areas. @ding2010modeling modelled
the spatial diffusion of mobile telecommunications across the highest
level of administrative division in China (29 regions in total). Their
analysis indicated that socioeconomic characteristics are important
determinants of the timing, speed and the level of mobile diffusion
within China. @bakher2013diffusion compared different the growth curves
as a means to model the diffusion of online shopping in Australia during
the 1998-2009 period, but they did not consider any spatial mechanisms.
At a global scale @PAPAGIANNIDIS2015308 modelled the diffusion of
different web technologies and practices across countries. Using similar
but less extensive data as the one used here, their analysis illustrated
how the diffusion of different web technologies and practices follows an
S-shaped pattern as well as the different diffusion rates of the
different technologies and practices. The most detailed consideration of
the spatial footprint of the diffusion of a new technology comes from
@lengyel2020role, who analysed the adoption and churn of a Hungarian
online social network. Their results were in agreement with the early
theoretical and empirical contributions reviewed here: assortativity,
urban scaling and distance are the key drivers of spatial diffusion.
Based on the above, the main empirical hypothesis states:

> H~1~: The three diffusion mechanisms namely distance, urban hierarchy
and the the S-shaped pattern of the cummulative level of adoption shaped
the diffusion of the commercial Web in the UK even at granular geographical 
scales.

@@

> H~2~: The diffusion of the commecrial Web is shaped by both early adoption 
and late comer effects.

These are the two hypotheses are empirically tested in @sec-results using the 
data and methods described in the next section.

@@

Another data cleaning step deals with some extreme outliers. Figure
\ref{correct} plots the website counts for the top 1000 postcodes. Some
obvious outliers can be observed for the 2002-2006 period for a handful
of postcodes, which can be attributed to link farms [@BL2013links]. The
website counts for these postcodes (SE24 9HP, CV8 2ED, GL16 7YA, CW1
6GL, M28 2SL, DE21 7BF), which in 2004 or 2005 had more than 1000
websites pointing to them, were replaced with predicted values based on
a simple panel regression model with postcode fixed effects and yearly
dummy variables. These postcodes refer to a residential area with a
small park in South London, residential areas outside Warwick and
Manchester, a rural area in the south border of England and Wales and
small business parks in Crew and Derby. Hence, it is difficult to
believe that these postcodes genuinely hosted such a large amounts of
websites, which were later extensively declined. To put the magnitude of
the data imputation into perspective, this process affected 6 out of the
557,808 postcodes included in the data. As the @sec-results illustrates,
this led to a small increase of the model predictive power.

@@

<!-- system-level analysis -->
These website density metrics are used for the three different stages of
the analysis. Firstly, a system-level analysis explores how the
diffusion of web technologies in the UK fits with the well-established
S-curve of cumulative adoption. To do so, the following logistic
function (Equation \ref{eq:s}) is estimated for the whole of the UK and
for each LAD separately:

@@

\noindent $k$ is the asymptote or, in other words, the saturation level,
$b$ the overall growth rate, and $t_{0}$ the *inflection point* of
maximum growth at $k/2$, where the logistic function is symmetrical
[@wilson201281]. The $t_0$ of each LAD is compared against the $t_0$ of
the UK to delineate whether a LAD reached that point faster or slower
than the country average. Importantly, an accuracy criterion was imposed
and only LAD with $R^2 > 0.9$ were included in this analysis. To
estimate Equation \ref{eq:s} a self-starting logistic growth model was
employed using the `nls` and `SSlogis` functions in `R`. This logistic
growth model depicts the initial period of slow diffusion of a new
technology, which is followed by the fast, exponential growth stage to
then eventually slow down and saturate [@wilson201281;
@grubler1999dynamics].

<!-- **minimum R2 = 95%** in @wilson201281, see also @grubler1990rise -->

<!-- The literature usually uses the saturation level as the asymptote. I am -->

<!-- using the total number of websites as we cannot compute a rate. -->

The system-level analysis also focuses on the volatility of the web
adoption to depict places with high concentration of early adopters and,
equally, latecomers. To do so, the change of the ranking of the UK LAD
over time is plotted and discussed. Both the S-curve and the volatility
analysis focus only on LAD as the very large number of OA would have
made such analysis difficult to visualise and interpret.

<!-- ESDA -->
Next, exploratory spatial data analysis depicts whether the two main
drivers of spatial diffusion -- namely neighbourhood and hierarchy --
underpin the diffusion of the Web in the UK. To capture the former and
following @ding2010modeling the Moran's I and the Local Index of Spatial
Association (LISA) are estimated for the website density. To address the
hierarchy mechanism, the Gini coefficient -- an established metric of
inequality -- is calculated. All the above are computed and plotted
longitudinally both for the LAD and the OA.

<!-- RF -->
Lastly, a modelling framework is developed to test the above diffusion
mechanisms. The overarching aim is to build a model that can test how
well these mechanisms can predict website density:

\begin{align}
Website\,Density_{t} \sim Hierarchy_{t-1} + Neighbourhood_{t-1} + S-curve_{t}\label{eq:rf.generic}
\end{align}

\noindent Equation \ref{eq:rf.generic} is estimated using Random Forest
(RF). This is a popular ML algorithm for both regression and
classification problems [@biau2012analysis]. It was introduced by
@breiman2001random and has become a go-to data science tool. RF can
effectively handle skewed distributions and outliers, model non-linear
relationships, require minimal hyperparameter tuning, exhibit low
sensitivity to these parameters, and have relatively short training
times [@Caruana2008; @liaw2002classification; @yan2020using]. These
attributes match well with the website density data characteristics
including skewness especially for the OA. Also, the large data size (c.
230k data points for each of the 17 years) calls for fast training
times. Importantly, RF predictions tend to be more accurate than those
from single regression trees and outperform Ordinary Least Squares in
out-of-sample predictions, even with moderate-sized training data and a
small number of predictors [@mullainathan2017machine; @athey2019machine;
@sulaiman2011intelligent; @pourebrahim2019trip; @biau2012analysis].

RF is a tree-based ensemble learning algorithm [@breiman2001random]. It
begins by generating random samples of the training data, which are then
used to grow regression trees to predict the dependent variable. Data
points and predictors are randomly sampled for the different trees. The
trees are trained in parallel using their own bootstrapped samples of
the training data. A crucial feature of RF is their ability not to
overfit, meaning they can generalize well to unseen test data. While
each tree may overfit individually, the ensemble of trees does not
because the errors of individual trees are averaged, reducing the
overall variance and preventing overfitting [@last2002improving]. For
regression problems, RF predictions are made by averaging the
predictions of all decision trees.

@@

## System-level analysis

<!-- S-shaped -->

```{r echo=FALSE, message=FALSE, fig.cap="\\label{s_uk}Grwoth curve, UK"}
path.image <- paste0(path, "/outputs/s/s_uk_per_firm.png")
knitr::include_graphics(path.image)
```

Figure \ref{s_uk} plots the cumulative adoption of the Web in total in
the UK during the 1996-2012 period. Not surprisingly, the cumulative
adoption is well represented by a curve that resembles the figure S. The
vertical line around year 2003 illustrates the point where the modelled
cumulative adoption is equal to 50% of the maximum. This $t_0$
*inflection* point signals the maximum adoption speed and is used here
to determine whether a UK LAD reached that inflection point earlier or
later than the UK average. Specifically, the S curve and the inflection
point are estimated for every LAD individually and then compared against
the UK average. LAD that reached their inflection point earlier than the
country average are labelled as *fast* and the rest as *slow*. Figure
\ref{s_map} maps this pattern. There is a clear high concentration of
fast adopting LAD in the South and East of England as expected. However,
there is also such a concentration in Wales, Scotland and the North West
of England. At a more detailed level, there are some expected examples
of LAD with relevant industrial backgrounds delineated as fast: the City
of London, a world-renowned cluster of finance industries
[@cook2007role], and Reading, a town with high-tech service industries
in proximity to London and its main airport, Heathrow
[@england2005polynet]. However, there are also LAD which were expected
to appear as fast -- e.g. Hackney in central London and Bristol, a
well-established creative cluster [@oatley1999cultural;
@bassett2002cultural] -- but were delineated as slow. Nevertheless,
there is obvious spatial clustering of fast and slow LAD. Importantly,
out of the ten fastest LAD, nine were located in the South East of
England and London and one in Cambridge (see Table \ref{table.s.lads} in
the Appendix). The above observations remain almost unchanged when the
analysis includes websites with up to 10 unique postcodes -- see Figures
\ref{s_uk10} and \ref{s_map10} in the Appendix -- advocating towards the
robustness of the analysis.

```{r echo=FALSE, message=FALSE, fig.cap="\\label{s_map}LAD adoption rates"}
path.image <- paste0(path, "/outputs/s/speed_map.png")
knitr::include_graphics(path.image)
```

<!-- discussion -->
The S-curves of the cumulative adoption of Web in the UK in total and in
the UK LAD demonstrate a pattern well aligned with all previous studies
discussed in @sec-litreview. Importantly, the above findings are in line
with previous studies exploring the diffusion of web technologies
[@PAPAGIANNIDIS2015308], social media [@lengyel2020role] and online
shopping [@bakher2013diffusion]. Similar to @beardsell1999spatial, who
analysed the spatial evolution of computer industry employment across
317 US metro areas, spatial heterogeneity is evident in the diffusion
speed when the scale of analysis is as detailed as the LAD. Importantly,
no previous studies estimated such cumulative adoption curves for areal
units as small as the UK LADS.

<!-- rankings -->
The next step assesses the stability and volatility of the LAD in terms
of their adoption of web technologies. As highlighted in the literature
[@risk_perceptions], different agents have different perceptions about
and levels of acceptance of the risks and the potential economic returns
associated with the adoption of new technologies -- see for instance the
seminal work of @venkatesh2000theoretical. To reveal such aggregated
patterns, Figure \ref{rank} plots the ranking of UK LAD based on website
density. To decrease noise, the average ranking of 1996-1998 and
2010-2012 is plotted instead of the individual years. The colours depict
the 5th and 95th percentile of the absolute difference of the LAD
ranking between 1996-1998 and 2010-2012. While there are quite a few
obvious cases of LAD that maintained their position between the
beginning and the end of the study period both at the top or at the
bottom of the hierarchy, there are also quite a few LAD that changed
drastically their position. Some of these LAD enjoyed a process that at
the first instance looks like *leapfrogging* since they managed to jump
at the top of the hierarchy despite their slow start. There is extensive
literature regarding the potential benefits of technological
leapfrogging. The underpinning argument is that latecomers can adopt and
benefit from new technologies that have been developed elsewhere without
incurring the hefty initial R&D costs [@teece2008firm]. Although the
leapfrogging literature does not pay much attention to cities and
regions [@yu2018sustainability], previous research highlighted the long
term and sustained productivity benefit of the early adopters of web
technologies in the UK [@tranosuk]. So, although the LAD system is
volatile, it is not clear whether the LAD with high concentration of
late adopters will gain any latecomer-related benefits in a way similar
to countries experiencing technological leapfrogging. Still, the
opportunity to catch up that @perkins2005international illustrated at
the country level, is also visible at the much more detailed LAD scale.
As previously, this is consistent with including websites with up to 10
postcodes (see Figure \ref{rank10}).

```{r echo=FALSE, message=FALSE, fig.cap="\\label{rank}Dynamics of wed diffusion (LAD)"}
path.image <- paste0(path, "/outputs/ranks/web_per_firm2000_2012_only0595_av.png")
knitr::include_graphics(path.image)
```

## Exploratory analysis

<!-- Neighbourhood effect -->
Figures \ref{morani} and \ref{lisa} offer a first insight into whether a
neighbourhood mechanism underpins the diffusion of the Web in the UK as
they present the Moran's I and the LISA maps of website density
respectively for LAD and OA. Starting from the former, spatial
autocorrelation was higher in the beginning of the study period and then
after 2000 dropped slightly and stabilised around 0.2. This reflects the
early concentration of high website density around London, which over
time diffused as high website density clusters can be seen in other
parts of the country away from London (Figure \ref{lisa}). An almost
reverse pattern can be observed for OA. At the beginning of the study
period Moran's I was around 0.5 and it plateaued after 2000 around 0.2.
Because of the very small size of OA, at the early stages of the
diffusion of web technologies their adoption was spatially scattered.
This is reflected in the lack of any significant clusters in 1996
(Figure \ref{lisa}). Eventually, as the adoption rate increased, more
such clusters of high website density were formed and this is reflected
both in the Moran's I and the LISA maps in Figures \ref{morani} and
\ref{lisa}. A similar pattern is observed for the extended dataset (see
Figures\ref{morani10} and \ref{lisa10}). In this case, the magnitude of
spatial autocorrelation is slightly higher illustrating a stronger
spatial clustering mechanism. All in all, the exploratory spatial data
analysis advocates towards an underpinning neighbourhood mechanism and
the different scales of analysis illustrate how it evolved differently
over time. <!-- discussion --> Interestingly, the magnitude of Moran's I
is similar to the ones revealed by @ding2010modeling about mobile phone
adoption in China albeit the much more detailed spatial scale of this
paper.

@@

To illustrate whether a hierarchical process also underpins the
diffusion of web technologies, the Gini coefficient is calculated yearly
both for LAD and OA. As a metric of inequality, the Gini coefficient
demonstrates whether website density is concentrated in a small number
of LAD or OA, or whether it is more equally spread across the country.
Both scales of analysis in Figure \ref{gini} illustrate the same
picture. At the beginning of the commercial Internet, website density
was extremely unequal, or, in other words, only a few places had
websites associated with them. Inequality dropped and plateaued after
2000 for both scales. This is illustrative of a hierarchical diffusion
mechanism that led over time to a more equal spread. Interestingly, the
year 2000 is again a period of change for this diffusion mechanism as it
was for the neighbourhood process. There is a substantial difference
between the Gini coefficient magnitude for LAD and OA, but this is
expected as the very small size of OA equates to a lot of polygons
without any websites pointing to them -- for example residential OA.

@@

## Modelling outputs

The next section incorporates the above mechanisms of the spatial
diffusion of the Web into a modelling framework. The aim is to use
variables depicting these spatial processes in order to predict the
diffusion of the Web as a new technology in the UK over space and time
and across different scales. Specifically, four different models are
estimated. Firstly, all the data points for the OA and LAD are utilised
in order to build two RF models and assess their capacity to predict the
adoption of the Web at the two different scales. These two models will
reveal the predictive capacity of the diffusion mechanisms and also
demonstrate how the importance of such variables changes across
different scales. The next two sets of models are again trained at these
two scales: OA and LAD. However, instead of using all the data points,
the OA and the LAD from one of the twelve UK regions are held out for
the model training. Then, the trained model is used to predict website
density for the OA or the LAD of the held-out region. This process takes
place recursively for all twelve UK regions. The regional differences in
the predictive capacity of the different samples will reveal how
dissimilar these spatial process are across regions and, importantly, at
different scales.

It needs to be highlighted here that the cross validation (CV) for all
models is spatially and temporally sensitive. Instead of using 10 random
and space- and time-agnostic folds,the `CAST` package is employed which
allows holding back data points from specific years and spatial units
and use them for testing in order to estimate the model performance
[@meyer2018improving].

>>>>

---
title: "A multi-scale story of the diffusion of a new technology: the web"
author:
  - name: Emmanouil Tranos
    email: e.tranos@bristol.ac.uk
    affiliations: 
      - id: some-tech
        name: University of Bristol
        department: 
        address: 
        city: 
        state: State
        postal-code: UK
    corresponding: e.tranos@bristol.ac.uk
address:
  - attributes:
        corresponding: true
    note: This is the first author footnote.
abstract: |
    This paper investigates the spatial diffusion of a new technology that is the Web in the UK. It employs novel data and machine learning methods to model the influence of well-established diffusion mechanisms. Contrary to previous studies, it adopts multiple scales, high spatial granularity and a long study period that captures the early stages of the Web until its maturity (1996-2012). Findings reveal the importance of such spatial mechanisms (namely distance, urban hierarchy and the S-shaped pattern of the cumulative level of adoption) even at granular scales. They also highlight spatial heterogeneity and instances of leapfrogging. 
keywords: 
  - diffusion
  - technologies
  - Web
  - digital
date: last-modified
bibliography: bibliography
format:
  elsevier-pdf:
    keep-tex: true
    text: |
      \usepackage[demo]{graphicx} % "demo" option just for this example
      \usepackage{subcaption}
    journal:
      name: Journal Name
      formatting: preprint
      model: 3p
      cite-style: authoryear
editor: 
  markdown: 
    wrap: 72
---

@@

Although the above processes have been studied in innovation studies,
the distancing of economic geographers from research questions about the
diffusion of new technologies and innovations resulted to a knowledge gap
regarding the explanatory power of these mechanisms at granular, local
scales [@ding2010modeling]. As discussed in @sec-introduction, such 
spatially explicit knowledge can directly inform local technology
and infrastructure policies. Indeed, while the spatial dimension is
present in the following studies, the geographical level of analysis is
often too coarse to ascertain the role of the above mechanisms at the
local scale. @beardsell1999spatial studied the evolution of the computer
industry in 317 US metropolitan areas during the 1977-1992 period using
employment data. Their analysis indicated that the relative size
distribution holds for urban computer employment and also urban
heterogeneity is essential in explaining this distribution. In a recent
study, @bednarz2020pulled focused on wind turbines and modelled their
spatial diffusion across 402 German regions during 1970-2015. Their key
finding is that local demand than local supply was the main driving
factor. @haller2011determinants employed firm panel data for Ireland
during 2003-2005 and illustrated that when firms are located in a region
or industry with high share of firms having a website, they are more
likely to have a website too. At a global scale @perkins2011internet
explored whether the adoption of previous communication technologies
that is mail, telegrams and telephones were shaped by similar
socioeconomic factors as the Internet. Their results indicated common
patterns regarding the drivers behind the adoption of different
communication technologies. @feldman2015rknowledge focused on inventions
and used patent data to map how the the rDNA technology diffused among
inventors across 366 US metropolitan areas. @ding2010modeling modelled
the spatial diffusion of mobile telecommunications across the highest
level of administrative division in China (29 regions in total). Their
analysis indicated that socioeconomic characteristics are important
determinants of the timing, speed and the level of mobile diffusion
within China. @bakher2013diffusion compared different growth curves
as a means to model the diffusion of online shopping in Australia during
the 1998-2009 period, but they did not consider any spatial mechanisms.
At a global scale @PAPAGIANNIDIS2015308 modelled the diffusion of
different web technologies and practices across countries. Using similar
but less extensive data as the one used here, their analysis illustrated
how the diffusion of different web technologies and practices follows an
S-shaped pattern as well as the different diffusion rates of the
different technologies and practices. The most detailed consideration of
the spatial footprint of the diffusion of a new technology comes from
@lengyel2020role, who analysed the adoption and churn of a Hungarian
online social network. Their results were in agreement with the early
theoretical and empirical contributions reviewed here: assortativity,
urban scaling and distance are the key drivers of spatial diffusion.
Based on the above, the main empirical hypothesis that is tested in this paper
states:

> H~1~: The three diffusion mechanisms namely distance, urban hierarchy
and the S-shaped pattern of the cumulative level of adoption shaped
the diffusion of the commercial Web in the UK even at granular geographical 
scales.

@@

> H~2~: The diffusion of the commercial Web is shaped by both early adoption 
and latecomer effects.

These two hypotheses are empirically tested in @sec-results employing 
the data and methods described in the next section.

@@

Another data cleaning step deals with some extreme outliers. Figure
\ref{correct} plots the website counts for the top 1000 postcodes. Some
obvious outliers can be observed for the 2002-2006 period for a handful
of postcodes, which can be attributed to link farms [@BL2013links]. The
website counts for these postcodes (SE24 9HP, CV8 2ED, GL16 7YA, CW1
6GL, M28 2SL, DE21 7BF), which in 2004 or 2005 had more than 1,000
websites pointing to them, were replaced with predicted values based on
a simple panel regression model with postcode fixed effects and yearly
dummy variables. These postcodes refer to a residential area with a
small park in South London, residential areas outside Warwick and
Manchester, a rural area in the south border of England and Wales and
small business parks in Crew and Derby. Hence, it is difficult to
believe that these postcodes genuinely hosted such a large amounts of
websites, which were later extensively declined. To put the magnitude of
the data imputation into perspective, this process affected 6 out of the
557,808 postcodes included in the data. As @sec-results illustrates,
this led to a small increase of the model predictive power.

@@

<!-- S -->
These website density metrics are used for the different stages of the analysis. 
Starting with the first hypothesis (H~1~), to test whether the S-curve 
accurately describes the cumulative adoption of the Web at a local scale, the
following logistic function (Equation \ref{eq:s}) is estimated for the whole of 
the UK and then for each LAD separately:

@@

\noindent $k$ is the asymptote or, in other words, the saturation level,
$b$ the overall growth rate, and $t_{0}$ the *inflection point* of
maximum growth at $k/2$, where the logistic function is symmetrical
[@wilson201281]. The $t_0$ of each LAD is compared against the $t_0$ of
the UK to delineate whether a LAD reached that point faster or slower
than the country average. This heterogeneity is discussed further when testing 
the second hypothesis (H~2~). Importantly, an accuracy criterion was imposed
and only LAD with $R^2 > 0.9$ were included in this analysis. To
estimate Equation \ref{eq:s} a self-starting logistic growth model was
employed using the `nls` and `SSlogis` functions in `R`. This logistic
growth model depicts the initial period of slow diffusion of a new
technology, which is followed by the fast, exponential growth stage to
then eventually slow down and saturate [@wilson201281;
@grubler1999dynamics].

<!-- **minimum R2 = 95%** in @wilson201281, see also @grubler1990rise -->
<!-- The literature usually uses the saturation level as the asymptote. I am -->
<!-- using the total number of websites as we cannot compute a rate. -->

<!-- ESDA -->
Next, exploratory spatial data analysis provides some first insights of whether 
the other two drivers of spatial diffusion -- namely neighbourhood and 
hierarchy -- underpin the diffusion of the Web in the UK. To capture the former 
and following @ding2010modeling the Moran's I and the Local Index of Spatial
Association (LISA) are estimated for the website density. To capture the
hierarchy mechanism, the Gini coefficient -- an established metric of
inequality -- is calculated. All the above are computed and plotted
longitudinally both for the LAD and the OA.

<!-- conceptual model -->
To formally test H~1~ a modelling framework is developed to assess the 
above diffusion mechanisms. The overarching aim is to build a model that can 
test how well these three mechanisms can predict website density:

\begin{align}
Website\,Density_{t} \sim Hierarchy_{t-1} + Neighbourhood_{t-1} + S\text-curve_{t}\label{eq:rf.generic}
\end{align}

<!-- <!-- conceptual model --> 
<!-- The models include variables that capture the three processes that the -->
<!-- relevant literature and the descriptive analysis highlighted. Namely, -->
<!-- the models capture: (i) a hierarchical mechanism with diffusion running -->
<!-- from main centres to secondary ones, (ii) a neighborhood mechanism -->
<!-- according to which diffusion first hits nearby locations, and (iii) the -->
<!-- rather canonical pattern of diffusion over time as reflected in the -->
<!-- S-shaped pattern in the cumulative level of adoption. -->

<!-- detailed features -->
To capture the hierarchical mechanism the model include as predictors a
one year lag of website density in London, the largest city in the UK, and a
one year lag of the website density in the nearest city. Both variables clearly
illustrate the urban hierarchy in the UK. Especially for the OA models, a 
third predictor is added to capture hierarchy at a much more granular scale:
a one year lag of the website density in the the nearest retail centre [^4]. As
the literature illustrates, retail centres are key components of urban systems
[@dennis2002central; @jones2021reframing]. Due to the small sizes of the retail 
centres, the latter is only relevant for and included in the OA models and not 
in the LAD models. In addition, the model includes the distance to London, the 
nearest city and the nearest retail centre. The underlying logic is that the 
level of website adoption in a spatial unit depends on the level of the 
adoption in places further up in the urban hierarchy the previous year. The
inclusion of the distance variables incorporates spatial structure into
the hierarchy argument. To depict the neighbourhood mechanism, the
website density of the neighbouring spatial units in the previous year
is employed. This is calculated using rook continuity. Again, the
underpinning rationale is that the level of web adoption within a
spatial unit depends on the level of web adoption in the neighbouring
spatial units the year before. This represents the 'hitting nearby
locations first' argument. Therefore, the spatial and temporal lag of
the website density in LAD and OA is calculated. Lastly, the temporal
pattern of the cumulative adoption is captured by a time trend variable.
Hence, the model will follow the following generic form (Eq.
\ref{eq:model}):

[^4]: The retail centres data have been obtained by the Consumer Data Research Centre, an ESRC data investment, under project ID CDRC 498-01, ES/L011840/1; ES/L011891/1: <https://data.cdrc.ac.uk/dataset/retail-centre-boundaries-and-open-indicators>.

\begin{align} \label{eq:model}
Website\,Density_{t} \sim Distance\,London +
Website\,density\,London_{t-1} +\notag\\
Distance\,Nearest\,City +
Website\,density\,Nearest\,City_{t-1} +\notag\\
Distance\,Nearest\,Retail_{i} +
Website\,density\,Nearest\,Retail_{t-1} +\notag\\
W*\, Website\,density_{t-1} +\notag\\ 
year_{t}
\end{align}

<!-- 4 models -->
Specifically, four different models are estimated. Firstly, all the data points 
for the OA and LAD are utilised in order to build two models and assess their
capacity to predict the adoption of the Web at the two different scales. These 
two models will reveal the predictive capacity of the diffusion mechanisms and 
also demonstrate how the importance of such variables changes across
different scales. The next two sets of models are again trained at these
two scales: OA and LAD. However, instead of using all the data points,
the OA and the LAD from one of the twelve UK regions are held out for
the model training. Then, the trained model is used to predict website
density for the OA or the LAD of the held-out region. This process takes
place recursively for all twelve UK regions. The regional differences in
the predictive capacity of the different samples will reveal how
dissimilar these spatial process are across regions and, importantly, at
different scales.

<!-- RF -->
Equation \ref{eq:model} is estimated using Random Forest
(RF). This is a popular ML algorithm for both regression and
classification problems [@biau2012analysis]. It was introduced by
@breiman2001random and has become a go-to data science tool. RF can
effectively handle skewed distributions and outliers, model non-linear
relationships, require minimal hyperparameter tuning, exhibit low
sensitivity to these parameters, and have relatively short training
times [@Caruana2008; @liaw2002classification; @yan2020using]. These
attributes match well with the website density data characteristics
including skewness especially for the OA. Also, the large data size (c.
230k data points for each of the 17 years) calls for fast training
times. Importantly, RF predictions tend to be more accurate than those
from single regression trees and outperform Ordinary Least Squares in
out-of-sample predictions, even with moderate-sized training data and a
small number of predictors [@mullainathan2017machine; @athey2019machine;
@sulaiman2011intelligent; @pourebrahim2019trip; @biau2012analysis].

<!-- RF -->
RF is a tree-based ensemble learning algorithm [@breiman2001random]. It
begins by generating random samples of the training data, which are then
used to grow regression trees to predict the dependent variable. Data
points and predictors are randomly sampled for the different trees. The
trees are trained in parallel using their own bootstrapped samples of
the training data. A crucial feature of RF is their ability not to
overfit, meaning they can generalize well to unseen test data. While
each tree may overfit individually, the ensemble of trees does not
because the errors of individual trees are averaged, reducing the
overall variance and preventing overfitting [@last2002improving]. For
regression problems, RF predictions are made by averaging the
predictions of all decision trees.

@@

## The three diffusion mechanisms at granular geographical scales

<!-- S-shaped -->
Figure \ref{s_uk} plots the cumulative adoption of the Web in the UK as a whole
during the 1996-2012 period. In line with the theoretical expectations discussed
in the previous section, the cumulative adoption of the Web is well represented 
by a curve that resembles the letter S. The vertical line in the beginning of 
2003 (2003.158) illustrates the point when the modelled cumulative adoption was 
equal to 50% of the maximum. This $t_0$ point is known as the *inflection* point 
and signals the maximum adoption speed. The results remain unchanged when the 
extended subset of websites, which include up to 10 unique postcodes is used
(see Figure \ref{s_uk10} in the Appendix). Importantly though, the extended
subset of websites indicates a saturation level close to 100% at the end of 
the study period.

```{r echo=FALSE, message=FALSE, fig.cap="\\label{s_uk}Cumulative adoption of the Web, UK"}
path.image <- paste0(path, "/outputs/s/s_uk_per_firm.png")
knitr::include_graphics(path.image)
```

What escaped from previous studies was εexploring whether such a pattern also 
describes cumulative adoption of new technologies at sub-national scales. To 
test this, the same  logistic function is estimated for each of the 374 LAD. The
detailed results are presented in Table \ref{table.s.lads} in the Appendix. 
The key finding is that for 342 LAD (91% of all LAD) the $R^2 > 0.9$ for the 
logistic function, which indicates that even at local scale the cumulative 
adoption of the Web is characterised by a pattern similar to the one observed 
for countries. Graphically, all these LAD curves are plotted in Figure 
\ref{s_lad_uk}. The different colour lines illustrate whether a LAD reached $t_0$ 
faster or slower than the UK in total. This heterogeneity is further discussed 
later in this section.

This is the first time that such cumulative adoption curves have been estimated
for areal units as small as the UK LAD. Understanding and modelling the dynamics
of technological diffusion and, ultimately, being able to predict diffusion 
rates at such small local scales can directly support the deployment of 
infrastructural networks (e.g. 5G) and local technology policies. The above are 
in line with previous studies exploring the diffusion of web technologies 
[@PAPAGIANNIDIS2015308], social media [@lengyel2020role] and online
shopping [@bakher2013diffusion].

```{r echo=FALSE, message=FALSE, fig.cap="\\label{s_lad_uk}Cumulative adoption of the Web, LAD"}
path.image <- paste0(path, "/outputs/s/lad_per_firm_all.png")
knitr::include_graphics(path.image)
```

<!-- Neighbourhood effect -->
To assess the second diffusion mechanisms, Figures \ref{morani} and \ref{lisa} 
offer a first insight into whether a neighbourhood or, in other words, a 
distance mechanism underpins the diffusion of the Web in the UK as they present 
the Moran's I and the LISA maps of website density respectively for LAD and OA.
Starting from the former, spatial autocorrelation was higher in the beginning of 
the study period and then after 2000 dropped slightly and stabilised around 0.2. 
This reflects the
early concentration of high website density around London, which over
time diffused as high website density clusters can be seen in other
parts of the country away from London (Figure \ref{lisa}). An almost
reverse pattern can be observed for OA. At the beginning of the study
period Moran's I was around 0.5 and it plateaued after 2000 around 0.2.
Because of the very small size of OA, at the early stages of the
diffusion of web technologies their adoption was spatially scattered.
This is reflected in the lack of any significant clusters in 1996
(Figure \ref{lisa}). Eventually, as the adoption rate increased, more
such clusters of high website density were formed and this is reflected
both in the Moran's I and the LISA maps in Figures \ref{morani} and
\ref{lisa}. A similar pattern is observed for the extended dataset (see
Figures\ref{morani10} and \ref{lisa10}). In this case, the magnitude of
spatial autocorrelation is slightly higher illustrating a stronger
spatial clustering mechanism. All in all, the exploratory spatial data
analysis advocates towards an underpinning neighbourhood mechanism, which 
is present at different scales. Interestingly, the magnitude of Moran's I
is similar to the ones revealed by @ding2010modeling about mobile phone
adoption in China albeit the much more detailed spatial scale of this
paper.

@@

To illustrate whether a hierarchical process also underpins the
diffusion of web technologies -- the third diffusion mechanism that is tested 
here -- the Gini coefficient is calculated yearly
both for LAD and OA. As a metric of inequality, the Gini coefficient
demonstrates whether website density is concentrated in a small number
of LAD or OA, or whether it is more equally spread across the country.
Both scales of analysis in Figure \ref{gini} illustrate the same
picture. At the beginning of the commercial Internet, website density
was extremely unequal, or, in other words, only a few places had
websites associated with them. Inequality dropped and plateaued after
2000 for both scales. This is illustrative of a hierarchical diffusion
mechanism that led over time to a more equal spread. Interestingly, the
year 2000 is again a period of change for this diffusion mechanism as it
was for the neighbourhood process. There is a substantial difference
between the Gini coefficient magnitude for LAD and OA, but this is
expected as the very small size of OA equates to a lot of polygons
without any websites pointing to them -- for example residential OA.

@@

After discussing the descriptive findings, the validity of the above diffusion
mechanisms at the local scale is assessed with a predictive modelling framework
described in @sec-datamethods. The reader is reminded that four models are 
estimated: for the OA and LAD, firstly all data points are employed to 
assess the capacity of the three mechanisms to to predict the adoption of the 
Web at these two scales; then, the LAD/OA from one of the twelve regions are
recursively held out to assess the spatial heterogeneity of these mechanisms
across the different UK regions.
